
---
title: "MLB11 Linear Regression Analysis"
author: Shreyu Umapathy
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
mlb11 <- read.csv("mlb11.csv")
```

## Batting average vs Runs

```{r batavg_model}
ggplot(mlb11, aes(x = bat_avg, y = runs)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Runs vs Batting Average", x = "Batting Average", y = "Runs")

batavg_model <- lm(runs ~ bat_avg, data = mlb11)
summary(batavg_model)
```

The scatterplot shows a positive linear relationship between batting average and runs. The linear model confirms this with a positive slope and high R².

## Compared to at_bats

```{r atbats_model}
atbats_model <- lm(runs ~ at_bats, data = mlb11)
summary(atbats_model)
```

Comparing R² values:

- **Batting average model R²**: `r round(summary(batavg_model)$r.squared, 3)`
- **At-bats model R²**: `r round(summary(atbats_model)$r.squared, 3)`

Batting average has a higher R², meaning it predicts runs better than at_bats.

## All traditional variables

```{r traditional_compare}
traditional_vars <- c("at_bats", "hits", "homeruns", "bat_avg", "rbi")
results <- data.frame(Variable = traditional_vars, R2 = NA)

for (i in seq_along(traditional_vars)) {
  var <- traditional_vars[i]
  model <- lm(as.formula(paste("runs ~", var)), data = mlb11)
  results$R2[i] <- summary(model)$r.squared
}

results <- results %>% arrange(desc(R2))
results
```

```{r best_traditional_plot}
best_trad_var <- results$Variable[1]
ggplot(mlb11, aes_string(x = best_trad_var, y = "runs")) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = paste("Runs vs", best_trad_var), x = best_trad_var, y = "Runs")
```

The best traditional predictor is `r results$Variable[1]`, based on the highest R² value.

## Moneyball variables

```{r moneyball_compare}
moneyball_vars <- c("obp", "slg", "ops")
moneyball_results <- data.frame(Variable = moneyball_vars, R2 = NA)

for (i in seq_along(moneyball_vars)) {
  var <- moneyball_vars[i]
  model <- lm(as.formula(paste("runs ~", var)), data = mlb11)
  moneyball_results$R2[i] <- summary(model)$r.squared
}

moneyball_results <- moneyball_results %>% arrange(desc(R2))
moneyball_results
```

```{r best_moneyball_plot}
best_moneyball_var <- moneyball_results$Variable[1]
ggplot(mlb11, aes_string(x = best_moneyball_var, y = "runs")) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = paste("Runs vs", best_moneyball_var), x = best_moneyball_var, y = "Runs")
```

The Moneyball variables mostly outperform traditional variables. `r moneyball_results$Variable[1]` has the highest R² and is the best predictor of runs.

## Model diagnostics

```{r diagnostics}
best_model <- lm(as.formula(paste("runs ~", best_moneyball_var)), data = mlb11)
par(mfrow = c(2, 2))
plot(best_model)
```

Model diagnostics for `r best_moneyball_var` show residuals are not too far off the line of best fit, This supports its validity as a strong linear predictor of runs.
